\section{Conclusion}
\label{sec:conclusion}

In this project we explored the implementation of a front-end for LISA. We argued that such a layered structure was a crucial design aspect to build a sound proof assistant. We demonstrated the feasibility and practicality of mixing backward and forward proof styles, and observed that it significantly helped users to write proofs. We introduced a concept of tactics, and we focused on the implementation of rules which provide a scalable interface to represent compound proof steps. We implemented a general matching procedure that can be used to infer most of the parameters when applying rules. We showed that our procedure was applicable to other use cases, such as the instantiation of justifications. On top of that, we also significantly improved the tooling which LISA can directly benefit from: a type safe DSL, a two-way parser/printer and various utilities to work with kernel proofs. Overall, our work was focused on interactivity and usability, and to that extent we have succeeded in producing a framework that was usable in a REPL environment. We exposed the limitations we have encountered or accidentally introduced, and provided ideas to solve them.

I would like to personally thank Viktor and Simon for giving me the opportunity to contribute early on to the design of LISA, and later for supervising this thesis. I would also like to thank all of the members of the LARA laboratory for the rich discussions and precious advice they shared.
